Mortality Notes

### 2014-11-18 proposal.Rmd
Wrote more intro, added template code for JAGS and Stan implementation of growth dependent and independent effects of wood density

Mortality rates vary both spatially and temporally, and as such, are critical to accurately predict vegetation dynamics. However, few models of vegetation dynamics explicitly account for changes in mortality rates. Instead, most models used simplifed assumptions such as constant mortality rates (Thorton 2007m White 2000, Delbart et al 2010) or that mortality varies as a function of some arbitary set of rules or thresholds (Sitch et al 2003, Sato et al 2007). In many cases, mortality is not simulated at all (Hawkes 2000). These simplifications are predominately a consequence of our limited understanding of what governs mortality rates and how it varies temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011). This limited knowledge is partly due to ecologists placing greater emphasis on examining plant growth and how it affects community and population dynamics (Hawke 2000). But also because of difficulties associated with modelling mortality (Hawke 2000). Mortality is highly variable and episodic (Burgmann 1994). As such, it requires individuals to be monitoried for sufficiently long periods of time, ideally from birth to death, which is problematic for long-lived species (maybe Bugman 1996). Mortality rates are also affected by multiple biotic and abiotic factors, as well as the frequency and severity of multiple hazards (e.g. fire, wind, drought), which may or may not be dependent on one another (McDowell 2011).

This complexity has led to the widespread use of phenomological models that attempt to describe patterns in plant mortality as a function of correlations with various covariates. In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic. Such phemenological models have been used to document how mortality varies as a function of competition (REF), climate (REF), age (REF) and size (REF). The effect of size has been one

These mechanistic models all assumed that mortality occured when carbon supply was insufficient for growth (e.g. Friend et al. 1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997; Fisher et al 2010). As such, none incorporated differential mortality rates among species, or how different hazards may alter these rates. Phenomological model on the other hand, have attempted to explain mortality as a function of competitive interactions, abiotic conditions as well as size. One finding that appears to

And the potential variation between sizes, classes, ages etc. is so high that even a large sample will not provide the statistical significance required.

Mechanistic models are constructed with al- gorithms that describe biological processes, as opposed to an empirical data set or predefined relationship. Thus mechanistic algorithms require no calibration, and are inherently deterministic.

Furthermore, ecologists must also asertain when an individual died and what caused its death, the determination of which is inherently difficult in plants.

#How have people attempted modelling mortality?
- In order to capture the U-shaped mortality researchers have used a variety of models to capture broad patterns. These include running seperate models for small and large plants, the use of polynomials, or the use of non-parameteric approaches that do not assume an underlying curve.
- They have often constructed these models to determine the effects of competition, size or ...
- However, the problem with such approaches is that they do not attempt to explain the underlying mechanisms that result in U-shaped mortality.
- Some have suggested that certain hazards influence different parts of the U-shape curve. For example, competitive process influence mortality when individuals are small and less so when they are large. Whereas the oppose is true for events such as windthrow.
- Furthermore, the vast majority of these empirical models have been conducted for individual species, and thus are of limited use for more applying to applying to other species.

@@

1) Scour the literature on how traits and size influence mortality rates under common hazards
2) Build an integrated process-based model that incorporates multiple hazards that can be used to determine how the combination of size and traits governs instantenous mortality rates.
3) Apply model to multiple datasets.

>>>>

Plant mortality and the processes affecting it are integral to understanding vegetation dynamics. However, few vegetation dynamic models explicitly simulate mortality (Hawkes 2000), and those that have, do so using little process-based information. For example, many models that account for plant mortality by assuming mortality rates are constant (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (Sitch et al 2003, Sato et al 2007). These simplifications are a result of a lack of process-based information on what causes mortality rates to vary temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011). In part, this lack of process-based knowledge is a consequence of researchers placing greater emphasis on delimiting what influences plant growth (Hawkes 2000). But it is also due to the difficulties associated with modelling and estimating mortality rates.

Temporal patterns in plant mortality are highly variable and episodic (Burgmann 1994). Mortality rates also vary between species and are strongly influenced by multiple biotic and abiotic factors that may interact with one another (e.g. drought and herbivory: McDowell 2011). Because of this complexity, improving process-based understanding of mortality requires monitoring large numbers of species and individuals across entire life-cycles and across multiple biotic and abiotic gradients. More importantly, it requires the cause of death to be known for each individual. However, few, if any, data meet all these requirements. Rather, most available data comes from monitoring programs that are usually shorter than many plants lifespans (Bugmann 1996), and that focus on single species at single sites. Furthermore, because of the inherent difficulty of defining the cause of death, most monitoring programs do not delimit the cause of death for each individual.

Consequently, there are very few mechanistic mortality models. In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic and all these were concerned with carbon budgets. In these models mortality occurred when carbon supply was insufficient for growth (e.g. Friend et al. 1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997; Fisher et al 2010). By contrast, the remaining 57 models were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997). These empirical mortality models could be either deterministic (REFS) or stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic processes such as competition (REF), climate (REF), age (REF), size (REF) and growth rate (REF).

More recently, researchers have used plant traits to determine why some species are more predispositioned to die under particular conditions than others.

One consistent finding among several phenomological models is that mortality rates tend to be U-shaped. That is, small and large plants are more prone to mortality relative to intermediate sized plants. However, the problem with these models is that most have been fitted for single species and thus may not accurately predict mortality in other species.

#How have people attempted modelling mortality?
- In order to capture the U-shaped mortality researchers have used a variety of models to capture broad patterns. These include running seperate models for small and large plants, the use of polynomials, or the use of non-parameteric approaches that do not assume an underlying curve.
- However, the problem with such approaches is that they do not attempt to explain the underlying mechanisms that result in U-shaped mortality.
- Some have suggested that certain hazards influence different parts of the U-shape curve. For example, competitive process influence mortality when individuals are small and less so when they are large. Whereas the oppose is true for events such as windthrow.
- Furthermore, the vast majority of these empirical models have been conducted for individual species, and thus are of limited use for more applying to applying to other species.

@@

1) Examine whether the effects of wood density are growth dependent or growth independent.
#Current JAGS model
```{r, eval=FALSE}
model <- function(x) {
  for(it in 1:n.obs) {

		died[it] ~ dbern(p[it])
		cloglog(p[it]) <- max(-9999, min(9999,  ln.census + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * dbh.gr[it]))))
	}

	#Priors
  cd0 ~ dlnorm(0, 0.0001)
  cd1 ~ dlnorm(0, 0.0001)
	cd2 ~ dlnorm(0, 0.0001)
	cd3 ~ dlnorm(0, 0.0001)

}
    test2<-     jags.parallel(model.file = model,
                 n.iter = 4000,
                 data= names(data), 
                 envir=list2env(data),
                 inits = NULL,
                 parameters.to.save = c('cd0','cd1','cd2','cd3'))
```

#Potential Stan Model
```{r, eval=FALSE}
library(parallel)
library(rstan)

cat('
  data {
		int<lower=0> N;
		int<lower=0, upper=1> M[N];
		vector[N] ln_T;
		vector[N] wd;
		vector[N] gr;

	}
	parameters {
		real<lower=0> c0;
		real c1;
		real<lower=0> c2;
		real<lower=0> c3;

	} 
	model {
		for (n in 1:N)
			M[n] ~ bernoulli(inv_cloglog(ln_T + log(c0 * exp(-c1 * wd[n]) + c2 * exp(-c3 * gr[n]))));
	}
', file=(StanModel <- tempfile())

BCImodel <- sflist2stanfit(
	mclapply(
		1:4,
		function(i) {
			stan(file=StanModel, data=BCIdata, seed=123, chains=1, chain_id=i, refresh=-1)
			}, 
		mc.cores=2
	)
)
```

2) Examine whether the effects of wood density varies with plant size.
3) Write a literature review summarising how plant traits influence mortality across multiple hazards (e.g. wind throw, drought, fire, disease, herbivory)
4) Build a process-based model based on what we have learnt from the above objectives.
